---
title: "DATA 607_ Week7_Homework (HTML XML JSON)"
author: "Jeremy O'Brien"
date: "March 16, 2018"
output: html_document
---

```{r setup, include = F}
knitr::opts_chunk$set(echo = TRUE)

library(XML)
library(RCurl)
library(stringr)
library(knitr)
library(kableExtra)
library(jsonlite)

options(knitr.table.format = "html")

setwd("G:/OneDrive/Learning/_CUNY_SPS_MSDS/2018_Spring/DATA 607/Week 7/Homework 7/Dataset/")
```

1. Take three files with identical data in HTML, XML, JSON.
2. Load data into three separate R data frames.
3. Review the three data frames and determine whether they're identical.

***

1-2. (HTML) Begin by collecting data from HTML file.

```{r, echo = T}
# The HTML file created is stored on github for ease of access.

html.url <- getURL("https://raw.githubusercontent.com/JeremyOBrien16/DATA-607/master/favbooks.html") 
data.html <- readHTMLTable(html.url, header = T, as.data.frame = T) 

# We give it a visual check.

data.html %>% kable

# Attempted to use kable_styling to improve look of resulting tables - strangely, nothing seemed to make any changes.
```

1-2. (XML) Next, collect data from XML file.

```{r, echo = T}
# Ditto, the XML file created is stored on github for ease of access.

xml.url <- getURL("https://raw.githubusercontent.com/JeremyOBrien16/DATA-607/master/favbooks.xml") 
file.xml <- xmlParse(file = xml.url)

# We call an XML function to get the data in a dataframe.

df.xml <- xmlToDataFrame(file.xml)

# We give it a visual check.

df.xml %>% kable()

# Attempted to use kable_styling to improve look of resulting tables - strangely, nothing seemed to make any changes.
```

1-2. (JSON) Lastly, collect data from JSON file.

```{r, echo = T}
# Likewise, the JSON file created is stored on github for ease of access.

json.url <- getURL("https://raw.githubusercontent.com/JeremyOBrien16/DATA-607/master/favbooks.json")
file.json <- (file = json.url)

# We call jsonlite's primary extractive function to get data into R, and then coerce that into a data frame.

data.json <- fromJSON(file.json)
df.json <- as.data.frame(data.json)

# We clean up the column headers.

colnames(df.json) <- str_extract_all(colnames(df.json), "(?<=\\.)[[:alpha:]]+")

# We give it a visual check.

df.json %>% kable()

# Attempted to use kable_styling to improve look of resulting tables - strangely, nothing seemed to make any changes.
```

***

3. Review the three data frames and determine whether they're identical.

```{r, echo = T}
# Examine the structure of each data frame.
str(data.html)
str(df.xml)
str(df.json)
```

***

* The HTML data is in a data frame within a list "NULL" with all variables as factors.
* The XML data is in a data frame with all variables as factors.
* The JXON data is in a data frame with variables as a mix of strings, numbers, and integers.

As a next step, we could harmonize each data.frame by coercing variables to the following classes:

* character strings: author[,1], title[,2], review[,7]
* factors: genre[,3], publisher_name [,4]
* numerics: price[,5]
* integers: publication_year[,6], pagenums[,8]

(Gave this a shot but it got messy so moved on.)
